Substitution rate response to changes in effective population size and protein expression level.

### 2020-06-16 manuscript/manuscript.tex
Manuscript formatted to MBE guidelines

\documentclass{article}
\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{caption}
\usepackage{mdframed}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage[margin=60pt]{geometry}
\pdfinclusioncopyfonts=1
\captionsetup{width=0.85\textwidth}
\renewcommand{\baselinestretch}{1.5}

\include{notations}
\begin{document}
	\title{Substitution rate responses to changes in effective population size.}
	\author{T. Latrille, N. Lartillot}
	\maketitle

	\abstract{
		The quasi-neutral theory of evolution asserts that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
		One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
		In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\omega$) with respect to $\Ne$.
		Such theoretical prediction had been observed in empirical data across many clades.
		However, several studies failed to observe such response of $\omega$ to changes in $\Ne$, or with weak strength or direction.
		Computational models of protein folding have observed that $\omega$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
		Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\omega$ and then a decrease.
		Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\omega$ responses to changes in $\Ne$.
		Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\omega$ in the context of a genotype-phenotype-fitness map.
		We derive general formulas, and specify an application in the case of fitness proportional to the probability of protein folding.
		Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
		We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\omega$ relationship in population genetic dataset.
		However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
		We also stress the importance of epistatic interactions in the $\Ne$-$\omega$ relationship, such that it determines to time to reach a new equilibrium.
	}

	\section*{Introduction}

	Molecular sequences differ across species due to the particular history of DNA substitutions along their respective lineages.
	These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
	These forces have effects at different levels: mutations are carried by molecular sequence, selection is mediated at the level of individuals, while random genetic drift is a population effect.
	Yet, they jointly contribute to the long-term molecular evolutionary process.
	Thus, the challenge of molecular evolution is to tease out their respective contributions, based on comparative analyses.

	One main aspect of this challenge is to correctly evaluate the role of random drift in the long term evolutionary process.
	Population genetics theory implies that the strength of drift, due to the stochastic sampling of mutations, is less pronounced in lineage with large effective population size ($\Ne$), and as a consequence, the purification by selection of weakly deleterious mutations is more effective.
	This fundamental idea is at the core of the nearly-neutral theory of evolution.
	The nearly-neutral theory posits that a substantial fraction of mutations are weakly deleterious.
	As a result, the theory predicts that the substitution rate of selected mutations relative to the neutral substitution rate, called $\omega$, decreases along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}.

	This prediction has been more quantitatively examined under the assumption that the selective effects of mutations are drawn from a fixed distribution of fitness effects (DFE) \cite{Kimura1979, Welch2008}.
	Assuming a gamma DFE, a key result obtained in this context is an approximate allometric scaling of $\omega$ as a function of $\Ne$ (i.e. $\omega \sim \Ne^{-\beta}$), where $\beta$ is the shape parameter of the DFE.
	In practice, DFEs are strongly leptokurtic, thus predicts a weak negative relation between $\omega$ and $\Ne$.

	The substitution rate of selected mutations relative to neutral substitution rate ($\omega$) is an observable quantity of the molecular evolutionary process, at least if we can separately estimate the rate of neutral and selected mutations in a phylogenetic or comparative context.
	Practically, in the case of protein-coding DNA sequences, and assuming that synonymous mutations are neutral, while those selected mutations are the non-synonymous changes affecting the amino-acid sequence, $\omega$ can be identified with the ratio of non-synonymous over synonymous substitution rates ($dN/dS$).
	Thus, in practice, the nearly-neutral argument translates into a predicted decrease in $\omega = dN/dS$ as a function of long-term changes in $\Ne$.

	The context of protein coding sequences fostered another modeling approach, based on genotype-fitness map instead of distribution of fitness effects.
	In this alternative approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998, Rodrigue2010, Tamuri2012}.
	For example, if the target amino-acid has a higher fitness than the current amino-acid, the selective effect of the mutation is positive, and reciprocally negative for a lower fitness.
	Such a modeling approach predicts higher substitution rates between amino-acid with similar biochemical properties.
	More fundamentally, the fitness depends solely on the current genotype, not on the trajectory of mutations leading to this sequence.
	Even though this modeling approach differs substantially from the one assuming a fixed DFE, it also predicts a negative correlation between $\omega$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.

	Empirically, variation in $\omega$ along the branches of phylogenetic trees has been inferred using phylogenetic codon models applied to empirical sequences \cite{Yang2001, Zhang2004}. 
	Combining estimations of $\omega$ along branches to life-history traits such as body mass or generation time revealed a positive correlation \cite{Popadin2007, Nikolaev2007}
	Subsequently, integrative inference methods combining molecular sequences and lineage specific quantitative traits have also found that $\omega$ correlates positively with traits such as longevity and body mass \cite{Lartillot2011, Figuet2017}.
	Since longevity and body mass typically correlate negatively with $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}, these empirical correlations suggest a negative correlation between $\omega$ and $\Ne$, thus confirming the theoretical prediction of the nearly-neutral theory of evolution.
	However, this correlation between $\omega$ and life-history traits could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation was found to be either not statistically significant (\cite{Lartillot2012}), or even in the opposite direction \cite{Lanfear2010, Nabholz2013, Weber2014, Figuet2016}.

	Mitigated empirical evidence about the response of $\omega$ to changes in $\Ne$ encouraged the search for alternative explanatory mechanisms \cite{Lanfear2014}.
	In this direction, one striking result was the proof that $\omega$ is in fact predicted to be independent of $\Ne$ under very general circumstances, namely, whenever (i) the fitness is a log-concave function of a phenotype and (ii) the phenotype itself is equimutable.
	Equimutability states that the distribution of phenotypic changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
	This general theoretical argument has been invoked in the context of \textit{in silico} experiments of protein sequence evolution, assuming that proteins are under selection for their thermodynamic stability, the fitness being proportional to the folding probability of the protein \cite{Goldstein2013}.
	The thermodynamic stability is itself computed using a 3D structural model of the protein. These simulation experiments have led to the observation that $\omega$ is essentially independent of $\Ne$.
	The explanation proposed for this result is that the distribution of changes in free energy of folding ($\deltadeltaG$) due to mutations is approximately independent of the current free energy ($\deltaG$), thus making the free energy of folding essentially equimutable.

	However, the equimutability assumption is a relatively strong one, which also conflicts with simple combinatorial considerations about the relation between sequence and phenotype. For example, if a protein sequence is already maximally stable, only destabilizing (or neutral) mutations can occur. More generally, assuming that the stability of a protein sequence reflects an underlying fraction of positions having already accepted destabilizing amino-acids, then the probability of destabilizing mutational events is in turn expected to directly depend on the current stability of the protein.

	In addition, if empirical evidence for a negative correlation of $\omega$ with $\Ne$ is still not totally convincing, another empirical correlation is known to be much more robust, namely, that of $\omega$ with expression level or protein abundance \cite{Duret2000, Rocha2004, Wang2011, Song2017}. Indeed, expression level is one of the best predictors of $\omega$, with highly expressed proteins typically having lower $\omega$ values. Of note, the slope of the response of $\omega$ to changes in expression level is relatively weak \cite{Drummond2005}, although clearly significant. Importantly, many of the theoretical models of protein-coding sequence evolution that have been invoked to explain this correlation between $\omega$ and expression level are such that the response of $\omega$ to changes in expression level or in $\Ne$ should be very similar. Conversely, under strict equimutability, most of these models would predict that $\omega$ should be essentially independent, not just of $\Ne$, but also of expression level, thus at odds with empirical observations.

	Overall, there is no doubt that the impact of changes in $\Ne$ (or in expression level) on $\omega$ is at best weak, making empirical estimation more difficult.
	Even if weak, however, an exact theoretical understanding and quantification of how the slope of this relation depends on the underlying map between genotype, phenotype and fitness, would be useful.
	Ultimately, relating proteins structural parameters to the $\omega$-$\Ne$ relationship reveal an intersection between disparate empirical dataset,  coming from protein thermodynamic on one side and comparative genomic on the other side.

	Lastly, the theoretical results discussed so far are all valid only when the balance between mutation, selection and drift is at equilibrium. However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ first leads to an increase of $\omega$ due to adaptive selection, and subsequently a decrease in $\omega$ due to stronger purifying selection in the long term \cite{Jones2016}.
	Studying only equilibrium properties can thus be misleading. For that reason, the dynamical response of $\omega$ to changes in $\Ne$ must also be addressed, quantified, and its connection with the underlying selective landscape better characterized.
	Dynamical properties of the $\omega$ response to changes of $\Ne$ are of theoretical interest, but are also empirically relevant such that if overlook they could thwart the relation between theoretical expectation and empirical estimates.

	% Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\omega$ relationship to $\Ne$ is unknown.
	The goal of this study is to characterize the dynamical and equilibrium response of $\omega$ to changes in $\Ne$, and relate this response to structural parameters of the model.
	To this end, we develop a general mathematical approach for deriving the $\omega$ relationship to $\Ne$ (and expression level) in the context of a given genotype-phenotype-fitness map.
	Based on these results, we discuss the consequences of the assumption that protein stability is a proxy of fitness, and the signatures persisting in DNA sequences observable under such assumption.
	In the light of previously published empirical estimates from protein thermodynamic, and from comparative genomics, we discuss the articulation between empirical data and our mechanistic model.
	We also discuss some of the alternative biophysical mechanisms that could determine the selective landscape on protein-coding sequences, and how they would modulate the slope of the relation between $\omega$ and $\Ne$.

@@

	The results that are presented below are valid for a general category of models of sequence evolution, based on an additive trait $x$ (i.e. such that the coding positions of the sequence contribute additively to the trait), under directional selection specified by a decreasing and log-concave fitness function $f(x)$. As a specific example, we more specifically consider a model of protein evolution under the constraint of thermodynamic stability (See figure \ref{fig:NeChangeInfluence}). This model is inspired from previous work \cite{Williams2006, Goldstein2011, Pollock2012}, except that we make several simplifying assumptions, allowing us to derive analytical equations. 
	%Throughout our derivation, we present the results both for the general case and for the more specific example.

	In the original biophysical model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
	Technically, the free energy of the folded are unfolded conformations are computed based on the $3$D conformations of the protein, and using a statistical potential.
	In such models, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.

	We approximate this model such that the destabilizing effect of an amino-acid does not depend on other neighbouring residues.
	Instead, each site contributes independently to $\deltaG$. 
	With such approximation, for each site of the sequence only one amino-acid is stabilizing the protein. 
	All $19$ other amino-acids are equally destabilizing, and each site contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ per destabilizing amino-acids.
	If all amino-acids of the sequence are stabilizing, $\deltaG$ is optimal and equal to $ \alpha < 0$. 
	In this model, the most succinct phenotype of a given genotype (i.e. sequence) is just the proportion of destabilizing amino-acids in the sequence, defined as $0 \leq x \leq 1$. Thus, $\deltaG$ is a linear function of $x$:
	\begin{align}
		\deltaG (x) = \alpha + \Nsite \gamma x,
	\end{align}
	where $\Nsite$ is the number of sites in the sequence.

	For a given $\deltaG$, thermodynamic equations allows to derive the proportion of protein in folded conformation in the cytoplasm, which is assumed to be a proxy for fitness.
	This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm (supp information).
	Analytically, this fitness by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
	\begin{align}
		f(x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma x)}},\\
		\Rightarrow f(x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma x)}, 
	\end{align}
	where $\alpha$ and $\gamma$ are defined as above, and the parameter $\beta$ is $1.686$ mol/kcal at room temperature.

@@

	In this section we present an analytical approximate solution for the response of equilibrium $\omega$ after a change in $\Ne$ (in log space). We call this reponse the susceptibility of $\omega$ to changes in $\Ne$, and denote it as $\chi$:
	\begin{align}
		\chi = \frac{ \der \omega}{\der \ln (\Ne)}
	\end{align}
	Broadly, deriving $\chi$ is done in two steps.
	The first step is to determine the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
	Subsequently, differential calculus on the response of the equilibrium phenotype to a change in $\Ne$ allows to ultimately derive an equation for $\chi$.
	The main results are given in the general case of any phenotype-fitness map, and also derived in the specific case where the fitness is proportional to the total amount of folded protein in the specific case of the biophysical model.
	All development are available in Supplementary Materials in the most general case.

	First, from a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of destabilizing amino-acid, or do nothing if the mutation is between both destabilizing amino-acids.
	To derive the probabilities of such events to occur, we also make the simplyfing assumption that the transition between amino-acids are equiprobable.
	Altogether, any mutation in the sequence can then have a phenotypic effect of $0$ or $\dx=\sfrac{1}{\Nsite}$, and the probabilities of transition are:
	\begin{gather}
		\begin{cases}
		\dx &\text{ with probability } 1-x, \\
		0 &\text{ with probability } \frac{18 x }{19}, \\
		-\dx &\text{ with probability } \frac{x}{19}.\\
		\end{cases} \label{eq:proba}
	\end{gather}
	In the extreme case of an optimal phenotype ($x = 0$), only destabilization mutations are proposed.
	Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or between destabilizing amino-acids (effect $0$) is proportional to $x$. 
	Or in other words, the mutation bias is proportional to $(1-x)/x$, which this is fundamentally a combinatorial effect, due to the number of mutational opportunities available in either direction.

	Second, we need to determine the strength of selection acting on mutations.
	Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
	\begin{align}
		s & \simeq \frac{1}{\Nsite}\frac{ \partial \ln f(x) }{\partial x} \label{eq:s} \\
		\Rightarrow s & \simeq - \beta \gamma e^{\beta (\alpha + \Nsite \gamma x)}. \label{eq:s-unfolded}
	\end{align}
	Conversely, stabilizing mutations will be under positive selection with opposite sign but same absolute value.
	It is important to realize that the selective effect is dependent on $x$, and that because fitnesses are log-concave, the absolute value of $s$ increases with regards to $x$.

>>>>

\documentclass{MBE}%
\usepackage[bookmarks]{hyperref}
\hypersetup{
	colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=black
}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\pdfinclusioncopyfonts=1
\include{notations}

\jshort{mbe}
\volname{}
\jvolume{}
\jvol{}
\jissue{}
\pubyear{}

\mstype{Article}

\begin{document}

\title[Substitution rate susceptibility] {Substitution rate response to changes in effective population size and protein expression level.}

\author[Latrille \& Lartillot] {
	T. \surname{Latrille}$^{\ast,1,2}$,
	N. \surname{Lartillot}$^{1}$
}

\address{
	$^{1}$Université de Lyon, Université Lyon 1, CNRS, Laboratoire de Biométrie et Biologie Évolutive UMR 5558, 69622 Villeurbanne, France.\\
	$^{2}$École normale supérieure de Lyon, Université de Lyon, Université Lyon 1, Lyon, France}

\coresp{thibault.latrille@ens-lyon.org or nicolas.lartillot@univ-lyon1.fr}

\abstract{
Molecular sequences are shaped by selection, where the strength of selection relative to drift is arbitrated by effective population size ($\Ne$). Populations with high $\Ne$ are expected to undergo stronger purifying selection, and consequently a decrease in substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\omega$=dN/dS). However, computational models based on biophysics of protein stability have suggested that $\omega$ can also be independent of $\Ne$, proven under general conditions. Together, the response of $\omega$ to changes in $\Ne$ depends on the specific mapping from sequence to fitness. Importantly,  increase in protein expression level has been found empirically to result in decrease of $\omega$, predicted by theoretical results recruiting the selection for protein stability. In this present work we derive a theoretical approximation for the response of $\omega$ to changes in $\Ne$ and expression level, under an explicit genotype-phenotype-fitness map. The method is generally valid for additive traits and log-concave fitness functions. We applied the results to protein undergoing selection for their conformational stability,  corroborated by simulations under more complex models. We predict a weak response of $\omega$ to changes in either $\Ne$ or expression level, which are interchangeable. Based on empirical data, we propose that fitness based on the conformational stability is not a sufficient mechanism to explain the amplitude of $\omega$ variations observed. However, other aspects of protein biophysics might be explored such as protein-protein interactions, which can lead to stronger response of $\omega$ to changes in $\Ne$.
}

\keyword{protein stability, substitution rate, population-genetics, drift, expression level.}

\maketitle

\section*{Introduction}
Molecular sequences differ across species due to the particular history of DNA substitutions along their respective lineages.
These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
These forces have effects at different levels: mutations are carried by molecular sequence, selection is mediated at the level of individuals, while random genetic drift is a population effect.
Yet, they jointly contribute to the long-term molecular evolutionary process.
Thus, the challenge of molecular evolution is to tease out their respective contributions, based on comparative analyses.

One main aspect of this challenge is to correctly evaluate the role of random drift in the long term evolutionary process.
Population genetics theory implies that the strength of drift, due to the stochastic sampling of mutations, is less pronounced in lineage with large effective population size ($\Ne$), and as a consequence, the purification by selection of weakly deleterious mutations is more effective.
This fundamental idea is at the core of the nearly-neutral theory of evolution.
This theory posits that a substantial fraction of substitutions are weakly deleterious, and as result predicts that the substitution rate of selected sites relative to the neutral substitution rate, called $\omega$, decreases along a lineage with higher $\Ne$ \citep{Ohta1972, Ohta1992}.

This prediction has been more quantitatively examined under the assumption that the selective effects of mutations are drawn from a fixed distribution of fitness effects (DFE) \citep{Kimura1979, Welch2008}.
Assuming a gamma DFE, a key result obtained in this context is an approximate allometric scaling of $\omega$ as a function of $\Ne$ (i.e. $\omega \sim \Ne^{-k}$), where $k$ is the shape parameter of the DFE.
In practice, DFEs are strongly leptokurtic, thus predicts a weak negative relation between $\omega$ and $\Ne$.

The substitution rate of selected mutations relative to neutral substitution rate ($\omega$) is an observable quantity of the molecular evolutionary process, at least if we can separately estimate the rate of neutral and selected mutations in a phylogenetic or comparative context.
Practically, in the case of protein-coding DNA sequences, and assuming that synonymous mutations are neutral, while non-synonymous changes affecting the amino-acid sequence are under selection, $\omega$ can be identified with the ratio of non-synonymous over synonymous substitution rates ($dN/dS$).
Thus the nearly-neutral argument translates into a predicted decrease in $\omega = dN/dS$ as a function of long-term changes in $\Ne$.

The context of protein-coding sequences fostered another modeling approach, based on genotype-fitness map instead of distribution of fitness effects.
In this alternative approach, the selective effect of a mutation depends on the fitness of both source and target amino-acids involved on the mutation \citep{Halpern1998, Rodrigue2010, Tamuri2012}.
For example, if the target amino-acid has a higher fitness than the current amino-acid, the selective effect of the mutation is positive, and reciprocally negative for a lower fitness.
Such a modeling approach predicts higher substitution rates between amino-acid with similar biochemical properties.
More fundamentally, the fitness depends solely on the current genotype, not on the trajectory of mutations leading to this sequence.
Even though this modeling approach differs substantially from the one assuming a fixed DFE, it also predicts a negative correlation between $\omega$ and $\Ne$ \citep{Spielman2015a, DosReis2015}.

Conversely, one striking theoretical result was the proof that $\omega$ is in fact predicted to be independent of $\Ne$ under very general circumstances, namely, whenever (i) the fitness is a log-concave function of a phenotype and (ii) the phenotype itself is equimutable.
Equimutability states that the distribution of phenotypic changes due to mutations is independent the current phenotype of the individual \citep{Cherry1998}.
This general theoretical argument has been invoked in the context of \textit{in silico} experiments of protein sequence evolution, assuming that proteins are under selection for their thermodynamic stability, the fitness being proportional to the folding probability of the protein \citep{Goldstein2013}.
The thermodynamic stability is itself computed using a 3D structural model of the protein. These computational experiments have led to the observation that $\omega$ is essentially independent of $\Ne$.
The explanation proposed for this result is that the distribution of changes in free energy of folding ($\deltadeltaG$) due to mutations is approximately independent of the current free energy ($\deltaG$), thus making the free energy of folding essentially equimutable.

However, the equimutability assumption is a relatively strong one, which also conflicts with combinatorial considerations about the relation between sequence and phenotype \citep{Serohijos2012}.
For example, if a protein sequence is already maximally stable, only destabilizing (or neutral) mutations can occur.
More generally, assuming that the stability of a protein sequence reflects an underlying fraction of positions having already accepted destabilizing amino-acids, then the probability of destabilizing mutational events is in turn expected to directly depend on the current stability of the protein.

Together, depending on the theoretical model mapping sequence to fitness, $\omega$ can be either independent or negatively correlated to $\Ne$, or even positively if considering adaptive evolution and environmental changes \citep{Lanfear2014}.
Empirically, variation in $\omega$ along the branches of phylogenetic trees has been inferred using phylogenetic codon models applied to empirical sequences \citep{Yang2001, Zhang2004}. 
Combining estimations of $\omega$ along branches to life-history traits such as body mass or generation time uncovered a positive correlation \citep{Popadin2007, Nikolaev2007}.
Subsequently, integrative inference methods combining molecular sequences and lineage specific quantitative traits have also found that $\omega$ correlates positively with traits such as longevity and body mass \citep{Lartillot2011, Figuet2017}.
Since lineage with a large body size and extended longevity typically correspond to low $\Ne$ \citep{Romiguier2014}, these empirical correlations suggest a negative correlation between $\omega$ and $\Ne$, thus confirming the theoretical prediction of the nearly-neutral theory of evolution.
However, the universality and robustness of the correlation between $\omega$ and life-history traits is still debated. 
Results have been not entirely consistent across independent studies, the correlation was found to be either not statistically significant \citep{Lartillot2012}, or even in the opposite direction depending on a specific clade or the potential biases taken into account \citep{Lanfear2010, Nabholz2013, Lanfear2014, Figuet2016}.

In addition, if empirical evidence for a negative correlation of $\omega$ with $\Ne$ is still not totally convincing, another empirical correlation is known to be much more robust.
Indeed, expression level or protein abundance is one of the best predictors of $\omega$, with highly expressed proteins typically having lower $\omega$ values, a correlation clearly significant although relatively weak \citep{Duret2000, Rocha2004, Drummond2005a, Zhang2015, Song2017}.
Importantly, theoretical models also based on protein stability have been invoked to explain this negative correlation between $\omega$ and expression level \citep{Wilke2006, Drummond2008}, such that selection against protein misfolding -due to toxicity- induces abundant proteins to evolve to greater stability, where the protein is more constrained and evolve slowly \citep{Serohijos2012}.

Indication that expression level and $\Ne$ play a similar roles on the evolution of protein have been noticed.
Precisely, under models of selection against protein misfolding, the free energy of folding $\deltaG$ varies similarly along a gradient of either $\Ne$ or expression level \citep{Serohijos2013}.
Additionally, under strict equimutability of $\deltaG$, computational models observed that $\omega$ is also independent of expression levels \citep{Serohijos2012}, similar to the phenomenon observed w.r.t $\Ne$.
However, the theoretical response of $\omega$ w.r.t to both $\Ne$ and expression level has not been quantified, and most importantly has not been related to the specific map between genotype, phenotype and fitness.
This analytical development would be useful to confront empirical data, relating $\omega$ to both $\Ne$ and expression level.
Ultimately, relating proteins structural parameters to response of $\omega$ reveals an articulation between empirical data from protein thermodynamic on one side and comparative genomics on the other side.

Lastly, the theoretical results discussed so far are all valid only when the balance between mutation, selection and drift is at equilibrium.
However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ first leads to an increase of $\omega$ due to adaptive selection, and subsequently a decrease in $\omega$ due to stronger purifying selection in the long term \citep{Jones2016}.
Studying only equilibrium properties can thus be misleading.
For that reason, the dynamic response of $\omega$ to changes in $\Ne$ must also be addressed, quantified, and its connection with the underlying selective landscape better characterized.
Dynamic properties of the $\omega$ response to changes of $\Ne$ are of theoretical interest, but are also empirically relevant such that if overlook they could thwart the relation between theoretical expectation and empirical estimates.
% Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\omega$ relationship to $\Ne$ is unknown.

This study aims to characterize the dynamic and equilibrium response of $\omega$ to changes in $\Ne$ and expression level, and relate this response to structural parameters of the model.
We develop a general mathematical approach to derive the $\omega$ response to $\Ne$ and expression level in the context of a given genotype-phenotype-fitness map.
In the light of previously published empirical estimates from protein thermodynamic and comparative genomics, we discuss the articulation between empirical data and our mechanistic model.
We also discuss some of the alternative biophysical mechanisms that could determine the selective landscape on protein-coding sequences, and how they would modulate the response of $\omega$ to changes in $\Ne$ and expression level.

@@

The results that are presented below are valid for a general category of models of sequence evolution, based on an additive trait $x$, such that the coding positions of the sequence contribute additively to the trait.
The trait is under directional selection specified by a decreasing and log-concave fitness function $f(x)$. As a specific example, we more specifically consider a model of protein evolution under the constraint of thermodynamic stability (See figure \ref{fig:Summary}, left panel). This model is inspired from previous work \citep{Williams2006, Goldstein2011, Pollock2012}, except that we make several simplifying assumptions, allowing us to derive analytical equations. 
%Throughout our derivation, we present the results both for the general case and for the more specific example.

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth, page=1] {artworks/summary.pdf}
\end{center}
 \caption{
 \textbf{Outline of the theoretical results.}
 Left panel. The complete state of the amino-acid sequence is the genotype. The phenotype ($x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of destabilizing sites in the sequence. The fitness is a decreasing log-concave function of the phenotype, depending on structural parameters of the model ($\Nsites, \DeltaGmin, \ \gamma, \ \beta$).
 Middle panel. Once defined the relation from genotype to fitness at the individual level, the evolutionary process consists of point substitutions of the amino-acid representative sequence of the population. Substitutions are mutation reaching fixation, thus the rate of substitution depends mechanistically on the interplay between mutation, selection and genetic drift.
 For a given effective population size ($\Ne$), the evolutionary process result in an average value of the phenotype ($\smash{x\eq}$) and an average relative substitution rate of selected mutation ($\omega$). Averaging over time is equivalent to determining the statistical equilibrium, by ergodicity of the stochastic process, given we can derive the probabilities of transitions between phenotypes.
 Right panel. The scaling of equilibrium $\omega$ under a range of $\Ne$ give us the susceptibility ($\chi$), as a function of the structural parameters defined in the phenotype-fitness map.
 }
 \label{fig:Summary}
\end{figure*}
In the original biophysical model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
Technically, the free energy of the folded and unfolded conformations are computed based on the $3$D conformations of the protein and statistical potentials.
In such models, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation, taking into account specific epistasis \citep{Dasmeh2018}.

We approximate this model such that the destabilizing effect of an amino-acid does not depend on other neighboring residues, disregarding specific epistasis \citep{Dasmeh2014}.
Instead, each site contributes independently to $\deltaG$. 
With such approximation, for each site of the sequence only one amino-acid is stabilizing the protein. 
All $19$ other amino-acids are equally destabilizing, and each site contribute to an excess of $\gamma = \Delta \Delta G > 0$ (in kcal/mol) to $\deltaG$ per destabilizing amino-acids.
If all amino-acids of the sequence are stabilizing, $\deltaG$ is optimal and equal to $ \DeltaGmin = \Delta G_{\text{min}} < 0$. 
In this model, the most succinct phenotype of a given genotype (i.e. sequence) is just the proportion of destabilizing amino-acids in the sequence, defined as $0 \leq x \leq 1$. Thus $\deltaG$ is a linear function of $x$:
\begin{align}
 \deltaG (x) = \DeltaGmin + \Nsite \gamma x,
\end{align}
where $\Nsite$ is the number of sites in the sequence.

For a given $\deltaG$, thermodynamic equations allows to derive the proportion of protein in folded conformation in the cytoplasm, which is assumed to be a proxy for fitness.
This fitness function is motivated in part that a protein must be folded to perform its function, however this a strong assumption which will be relaxed subsequently to take into account protein expression level.
Analytically, this fitness is given by Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation \citep{Goldstein2011}: 
\begin{align}
 f(x) = \dfrac{1}{1 + e^{\beta (\DeltaGmin + \Nsite \gamma x)}},\\
 \Rightarrow f(x) \simeq 1 - e^{\beta (\DeltaGmin + \Nsite \gamma x)}, 
\end{align}
where $\DeltaGmin$ and $\gamma$ are defined as above, and the parameter $\beta$ is $1.686$ mol/kcal at room temperature.

@@

In this section we present an analytical approximate solution for the response of equilibrium $\omega$ after a change in $\Ne$ (in log space). We call this response the susceptibility of $\omega$ to changes in $\Ne$, and denote it as $\chi$:
\begin{align}
 \chi = \frac{ \der \omega}{\der \ln (\Ne)} \label{eq:chi}
\end{align}
Broadly, deriving $\chi$ is done in two steps (See figure \ref{fig:Summary}).
First, we determine the phenotype at equilibrium, when evolutionary forces of mutation, selection and genetic drift compensate each other.
Subsequently, differential calculus on the response of the equilibrium phenotype to a change in $\Ne$ allows to ultimately derive an equation for $\chi$.
The main results are given in the general case of any phenotype-fitness map, and also derived in the specific case of the biophysical model, where the fitness is proportional to the total amount of folded protein.
All developments are available in supplementary materials in the most general case.

From a given genotype, mutations changing the sequence have various effects, they can increase or decrease the proportion of destabilizing amino-acid, or do nothing if the mutation is between two destabilizing amino-acids.
To derive the probabilities of such events to occur, we also make the simplifying assumption that all transitions between amino-acids are equiprobable.
Altogether, any mutation in the sequence can then have a phenotypic effect of $0$ or $\dx=\sfrac{1}{\Nsite}$, and the probabilities of transition are:
\begin{gather}
 \begin{cases}
 \dx &\text{ with probability } 1-x, \\
 0 &\text{ with probability } \frac{18 x }{19}, \\
 -\dx &\text{ with probability } \frac{x}{19}.\\
 \end{cases} \label{eq:proba}
\end{gather}
In the extreme case of an optimal phenotype ($x = 0$), only destabilizing mutations are proposed.
Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or in between destabilizing amino-acids (effect $0$) is proportional to $x$. 
Or in other words, the mutation bias is proportional to $(1-x)/x$, which this is fundamentally a combinatorial effect, due to the number of mutational opportunities available in either direction.

Second, we need to determine the strength of selection acting on mutations.
Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
\begin{align}
 s & \simeq \frac{1}{\Nsite}\frac{ \partial \ln f(x) }{\partial x} \label{eq:s} \\
 \Rightarrow s & \simeq - \beta \gamma e^{\beta (\DeltaGmin + \Nsite \gamma x)}. \label{eq:s-unfolded}
\end{align}
Conversely, stabilizing mutations will be under positive selection with opposite sign but same absolute value.
It is important to realize that the selective effect is dependent on $x$, and that because fitness are log-concave, the absolute value of $s$ increases with regards to $x$.

Based on these expressions for mutational and selective pressure, one can study the evolutionary process trajectory.
Starting from an optimal sequence, mostly destabilizing mutations will occur, but they may reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, at which point the protein will reach a point of equilibrium called marginal stability \citep{Taverna2002, Bloom2007}.
Most importantly the probability of fixation of mutations is affected by genetic drift, parameterized by effective population size ($\Ne$).
Altogether, at the equilibrium between mutation, selection and drift, the selection coefficient of new advantageous and deleterious mutations reaching fixations are expected to be null on average \citep{Goldstein2013}.
Formally, and after simplification, the equilibrium phenotype denoted $x\eq$ is given in the general case by the equation :
\begin{align}
 \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(x\eq) }{\partial {x\eq}} \\
 \Rightarrow \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq 4\Ne \beta \gamma e^{\beta (\DeltaGmin + \Nsite \gamma x\eq)} \text{,} \label{eq:equilibrium}
\end{align}
in the more specific case of the biophysical model. This equation essentially expresses the mutation-selection equilibrium: the left-hand side of the equation is the log of the mutation bias at $x$, while the right-hand side is simply $4 \Ne s$, the scaled selection coefficient.
